---
title: "How to use GGIRmatcher"
author: "Jairo H Migueles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use GGIRmatcher}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The purpose of `GGIRmatcher` is to match accelerometer data processed
with the [GGIR
package](https://cran.r-project.org/web/packages/GGIR/index.html) with
data collected from other sensors. In this first available version of
the package, all the procedures have been tested using continuous
glucose monitors, thus, further developments might be needed to adapt
`GGIRmatcher` to the characteristics and requirements of other types of
sensors.

The package is designed to be able to match the data from GGIR with the
data from the other sensor based on the timestamps, thus it is the user
responsibility to make sure that time stamps are comparable across
sensors as GGIRmatcher will trust the input time stamps.

## Pre-requisites

Using GGIRmatcher require that:

1.  Accelerometer data has been processed beforehand with GGIR (you are
    expected to provide the GGIR output directory as input parameter in
    GGIRmatcher).

2.  The additional sensor data has been conveniently cleaned and stored
    as RData files.

### Example of GGIR function call

The bare minimum of GGIR parameters that needs to be defined for GGIRmatcher to
work is as follows:

``` r
library(GGIR) 
GGIR(datadir = "C:/myproject/acc_files/",
     outputdir = "./",
     idloc = 2,
     save_ms5rawlevels = TRUE, 
     save_ms5raw_format = "RData",
     save_ms5raw_without_invalid = FALSE,
     timewindow = c("MM", "WW", "OO"))
```

### Example of additional sensor data cleaning

As any sensor implies different data cleaning necessities, this is a
step that each user should approach based on their needs and
requirements. The RData files from the additional sensor are expected to
only contain a two-column data frame, including: time stamp in ISO8601
format and the output of interest.

This package contain an example raw recording from a continous glucose
monitor that comes in the format of ".csv". Table 1 shows the first rows
of such a recording.

```{r rawfile, echo=FALSE, tab.cap="Table 1. First records from a raw continuous glucose monitor recording (accesible at 'inst/extdata/cgm_raw_sample.csv')."}
raw = data.table::fread(system.file("extdata", "cgm_raw_sample.csv", 
                                    package = "GGIRmatcher"))
knitr::kable(raw[1:4,1:5])
```

This file was read and cleaned so that a two-column data frame with time
stamp in ISO8601 format and glucose readings are stored in RData files.
Note that, conveniently, also the units of measurement for glucose were
converted to mg/dL in the cleaning process (Table 2).

```{r cleanfile, echo=FALSE, tab.cap="Table 2. Cleaned version of the raw continous glucose monitor data."}
knitr::kable(GGIRmatcher::cgm_cleaned[1:4,])
```

## How to match the data with GGIRmatcher

### 1- Match time series at epoch level

First, the function `match_time_series` can be used to load the time
series from GGIR and the time series from the other sensor. Time zones
and daylight saving time days are considered and handled conveniently in
this function.

``` r
library(GGIRmatcher) 
match_time_series(
  GGIR_output_dir = "C:/myproject/output_myproject",
  additional_ts_dir = "C:/myproject/output_myothersensor",
  outputdir = "C:/myproject/output_matched",
  add_metric_name = "GLUC",
  idloc = "_")
```

The description of the parameters in the function above can be found at
the documentation file for the function. In brief, these are the
parameters we are using in this example:

-   **GGIR_output_dir = "C:/myproject/output_myproject":** path to the
    folder containing the output produced by GGIR (in this example, this
    would be the folder "output_myproject")

-   **GGIR additional_ts_dir = "C:/myproject/output_myothersensor":**
    path to the folder containing the cleaned output data from the other
    sensor (in this example, the folder "output_myothersensor"). This
    would be, the RData files containing the data frames as the one
    shown in Table 2.

-   **outputdir = "C:/myproject/output_matched":** path to the folder
    that will be used by GGIRmatcher to store the new output produced
    (here, "output_matched").

-   **add_metric_name = "GLUC":** character with the column name in the
    cleaned output that contains the metric to be matched with the
    acceleration from GGIR. In this example, the name is GLUC as can be
    seen in Table 2. If this is not provided, GGIRmatcher will try to
    guess the name by looking for column names that are not supposed to
    be part of the GGIR output, yet this can be risky.

-   **idloc = "\_":** character used for the identificaiton of the
    recording ID. This is important as GGIRmatcher will match GGIR
    output with the additional sensor based on the ID. Therefore, the
    IDs should be exactly the same in the GGIR output and here.
    GGIRmatcher extracts the ID from the file name, and will use the
    part of the file name preceding the first occurrence of the
    character indicated in idloc.

The output of this function is a set of RData files (one per matched
recording) stored in the output directory (within the "meta/ms5.outraw")
subfolder. These RData files contain a data frame named ts with 13
columns.

```{r matchedfile, echo=FALSE, tab.cap="Table 3. Matched CGM and accelerometer data."}
matched_cl = GGIRmatcher::matched
matched_cl[is.na(matched_cl)] = ""
knitr::kable(matched_cl[1:4,])
```

In this case, the CGM monitor started collecting data before the accelerometer
was initialized. Therefore, the first rows do only contain CGM data. Figure 1
shows the CGM and accelerometer data matched.

```{r fig1, echo=FALSE, fig.width=8, fig.cap="Figure 1. Visualization of the first 2 days of data collected with the two devices."}
from = as.numeric(as.POSIXct("2024-04-01T12:00:00+0200", 
                             format = "%Y-%m-%dT%H:%M:%S%z"))
to = grep("2024-04-03T12:00:00", GGIRmatcher::matched$timestamp)
d = GGIRmatcher::matched[1:to,]
time = as.POSIXct(d$timestamp, format = "%Y-%m-%dT%H:%M:%S%z")
xlim = as.POSIXct(c("2024-04-01T12:00:00+0200",
                    d$timestamp[nrow(d)]),
                  format = "%Y-%m-%dT%H:%M:%S%z")
wu = which(diff(c(0,d$SleepPeriodTime)) == -1)
so = which(diff(c(0,d$SleepPeriodTime)) == 1)
wu = as.POSIXct(d$timestamp[wu], format = "%Y-%m-%dT%H:%M:%S%z")
so = as.POSIXct(d$timestamp[so], format = "%Y-%m-%dT%H:%M:%S%z")
nights = as.POSIXct(grep("00:00:00", d$timestamp, value = T), format = "%Y-%m-%dT%H:%M:%S%z")
# plot
par(mfrow = c(2,1), mar = c(1,4,1,1), las = 1)
plot(time, d$ACC, type = "l", lwd = 1, col = "#004D40",
     xlim = xlim, ylim = c(0, 300), axes = F, bty = "n",
     ylab = "", xlab = "")
title(ylab = "Acceleration (mg/dL)", cex.lab = 0.8)
rect(xleft = so, xright = wu, ybottom = -50, ytop = 300,
     col = "#D3E8FA", border = NA)
lines(x = time, y = d$ACC, col = "#004D40")
abline(v = seq.POSIXt(xlim[1], xlim[2], by = 4*60*60),
       col = "grey", lty = "dotted")
axis(side = 2, cex.axis = 0.8)
abline(v = nights)
par(mar = c(2,4,0,1))
sel = which(!is.na(d$GLUC))
plot(time[sel], d$GLUC[sel], type = "l", lwd = 2, col = "#D81B60",
     xlim = xlim, axes = F, bty = "n",
     ylab = "", xlab = "")
rect(xleft = so, xright = wu, ybottom = -5, ytop = 300,
     col = "#D3E8FA", border = NA)
title(ylab = "Glucose (mg/dL)", cex.lab = 0.8)
axis(side = 2, cex.axis = 0.8)
lines(x = time[sel], y = d$GLUC[sel], col = "#D81B60", lwd = 2)
axis(side = 1, at = seq.POSIXt(xlim[1], xlim[2], by = 4*60*60),
     labels = format(seq.POSIXt(xlim[1], xlim[2], by = 4*60*60), format = "%H"),
     tick = F, line = -0.5)
abline(v = nights)
par(mar = c(2,4,0,1))

```

### 2- Aggregate per window of interest

Next, the function aggregate_per_window aggregates the data in some different window
types, following the methods used by GGIR. This is:

-   **MM windows:** midnight to midnight, this matches the calendar
    dates.

-   **WW windows:** windows starting the moment the participant wakes up
    and finishing the moment the participant wakes up the next day.
    Wake-up times are the ones detected in GGIR.

-   **OO windows:** windows starting the moment the participant falls
    sleep and finishing the moment the participant falls sleep the next
    day. Sleep onset times are the ones detected in GGIR.

-   **Segments of the day:** User-defined segments of the day. As for
    now, GGIR only supports integer indicators of segments of the day,
    for example: c(0, 6, 12, 18, 24) will include windows from 0am to
    6am, from 6am to 12pm, from 12pm to 6pm, and from 6pm to the next
    0am.

For the data aggregation, the user should define how the data will be
aggregated, i.e., what functions should be applied to the additional
sensor data. This can be as simple as calculating the mean, or as
complex as including long functions with advanced calculations. In this
example, we are using a list of two functions to calculate the mean, and
the standard deviation for the additional sensor data:

``` r
myfuns = list(mean = function(x) mean(x, na.rm = T),  
              SD = function(x) sd(x, na.rm = T))
```

``` r
aggregate_per_window(tspath = "output_matched/meta/ms5.outraw/",
                     outputdir = "C:/myproject",
                     GGIR_output_dir = "C:/myproject/output_myproject",
                     qwindow = c(0, 6, 12, 18),
                     FUNs = myfuns, overwrite = F)
```

The output of this function is a set of RData files (one per matched
recording) stored in the output directory (within the "meta/ms5.out")
subfolder. These RData files contain a with all the window summaries 
that apply. Table 4 shows an example.

```{r aggperwindow, echo=FALSE, tab.cap="Table 4. Window aggregates for the example file."}
sel = which(GGIRmatcher::aggperwindow$calendar_date %in% c("2024-04-01",
                                                           "2024-04-02",
                                                           "2024-04-03") &
              GGIRmatcher::aggperwindow$window %in% c("MM", "OO", "WW"))
vars = c("ID", "window", "window_number", "calendar_date", "weekday",
         "start_date", "start_time", "end_date", "end_time",
         "dur_day_spt_min", "dur_day_min", "dur_spt_min",
         "dur_spt_sleep_min", "dur_day_total_IN_min", 
         "dur_day_total_LIG_min", "dur_day_total_MOD_min",
         "dur_day_total_VIG_min",
         "mean_GLUC_day_spt", "mean_GLUC_day", "mean_GLUC_spt",
         "SD_GLUC_day_spt", "SD_GLUC_day", "SD_GLUC_spt")
show = GGIRmatcher::aggperwindow[sel,vars]
knitr::kable(show[order(show$window),])
```

### 3- Extract full window-level report

Finally, the last step in the process is to store the reports of interest in csv format. `GGIRmatcher` will only store a report per window type, with an indicator of the valid windows for the accelerometer (GGIR output) and for the additional sensor (in this case, the CGM output). For this, we use the `full_window_report function`.

``` r
full_window_report(outputdir = "C:/myproject",
                   GGIR_output_dir = "C/myproject/output_myproject/",
                   additional_sf_seconds = 900,
                   min_WW_OO_length = 16,
                   min_segment_length_hours = 6
                   # GGIR cleaning parameters 
                   includedaycrit = 0.7, 
                   includenightcrit = 0.7, 
                   includedaycrit.part5 = 0.7,
                   # cleaning parameters for additional monitor
                   includecrit_day_spt = 0.7, 
                   includecrit_day = 0.7, 
                   includecrit_spt = 0.7)
```

## The output of GGIRmatcher

As probably noticed by now, `GGIRmatcher` mimics the output folder 
structure of GGIR and creates a meta and a results folder with the 
pertinent sub-folders:

**meta** folder

-   ms5.out

-   ms5.outraw

**results** folder

-   QC

    -   full window-level report (MM, WW, OO, Segments)

-   variableDictionary (user is responsible for filling up the
    additional sensor output descriptions)
