```{r}

```

---
title: "How to use GGIRmatcher"
author: "Jairo H Migueles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use GGIRmatcher}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GGIRmatcher)
```

# Pre-requisites

Using GGIRmatcher require that you have previously run GGIR over your accelerometer data, and additionally that you have cleaned and stored your additional sensor data as RData files.

As additional sensors might have different cleaning necessities, this is a step that each user should approach based on their needs and requirements. The RData files from the additional sensor are expected to contain a two-column data frame, including: timestamp and the output of interest.

Also, it is important that the RData files do only contain this data frame, with no additional objects. This package includes an example raw output file from a

```{r echo=FALSE}

```

# How to match the data

## 1- Match time series at epoch level

``` r
library(GGIRmatcher) 
match_time_series(
  GGIR_output_dir,
  additional_ts_dir,
  outputdir,
  add_metric_name = NA,
  idloc = "_",
  tz = Sys.timezone(),
  overwrite = F,
  verbose = T)
```

## 2- Aggregate per window

1.  Define functions to apply to the additional sensor data:

Here some examples.

``` r
myfuns = list(mean = function(x) mean(x, na.rm = T),        # mean
              median = function(x) median(x, na.rm = T),    # median
              SD = function(x) sd(x, na.rm = T),            # standard deviation
              CV = function(x) sd(x, na.rm = T) / mean(x, na.rm = T), # coefficient of variation
              nCGM_levels = function(x) table(cut(x, breaks = c(0,54,70,140,180,250,Inf), right = F)), # n recordings in levels of the additional sensor output
              AUC = function(x, time) pracma::trapz(time[!is.na(x)]/60, x[!is.na(x)])) # area under the curve
```

2.  Aggregate per window:

``` r
aggregate_per_window(
  tspath,
  outputdir,
  GGIR_output_dir,
  FUNs = list(n = function(x) sum(!is.na(x)), mean = mean),
  qwindow = NULL,
  qwindow_names = NULL,
  overwrite = FALSE,
  tz = Sys.timezone(),
  verbose = TRUE)
```

## 3- Extract full window-level report

``` r
full_window_report(
  outputdir,
  GGIR_output_dir = NULL,
  includecrit_day_spt = NULL,
  includecrit_day = NULL,
  includecrit_spt = NULL,
  additional_sf_seconds = NULL,
  minimum_WW_OO_length = 0,
  minimum_segments_length = 0,
  verbose = TRUE,
  ...)
```

# The output of GGIRmatcher

It mimics GGIR output folder structure

meta folder

-   ms5.out

-   ms5.outraw

results folder

-   QC

    -   full window-level report (MM, WW, OO, Segments)

-   variableDictionary (user is responsible for filling up the additional sensor output descriptions)
